Slitasje og kjørespor

### 2023-01-25 DRAFT_slitasjeOgKjørespor.Rmd
feat: spread to HIAs

This rather dense figure needs to be looked at carefully. 
In the top row we see that the sample size (number of polygons) varies between ecosystems. For naturally open ecosystems many polygons are from high infrastructure areas, not surprisingly, and there are few points for the lowest impact class (zero such polygins in _Sørlandet_ for example). We there may not be able to extrapolate indicator values in areas in _Sørlandet_ with low levels of infrastructure. We can also from the middle row see the association between the indicator values (area weighted means) and the modified infrastructure index (HIF). We should then not put too much weight on data were there is little data (low n) and/or high standard error (sd, bottom row). For example, with Naturally open ecosystems, Midt-Norge has the overall lowest errors, and there we also see the kind of relationship between the indicator and the pressure that we want to se (middle row). The pattern is repeated for all regions except Østlandet, but there we also see that HIF class 3 has very little data, and a high sd, and this could explain the discrepancy (but it doesn't support the general relationship either).

@@

@@

We can also calculate Pearsons correlation coefficients.

```{r}
corrCheck_grouped <- as.data.frame(corrCheck) %>%
  group_by(hovedøkosystem, region) %>%
  summarise(cor = cor(i_2020, infrastructureIndex))
```

@@

```{r, fig.cap="A closer look at the HIA designation over Trondheim"}
tm_shape(HIA_trd)+
  tm_polygons(col = "infrastructureIndex",
    title="Infrastructure index\n(modified 4-step scale)",
    palette = "viridis",
    style="cat")+
  tm_layout(legend.outside = T)+
  tm_shape(hw_utm)+
  tm_lines(col="blue")
```

@@

I want to add a threshold so that we don't end up over extrapolating
based on too few data points. I think we need a minimum of 100 data point to base an estimate on.

<!-- NEEDS CACHING -->
```{r}
mySummary <- ea_spread(indicator_data = wetlands,
                         indicator = i_2020,
                         regions = HIA_regions,
                         groups = region_HIF,
                         summarise = TRUE,
                       threshold = 100)
```

<!-- Do we need another aggregation step here, to get regional estimates= -->

Let's split the regions column

```{r}
mySummary <- mySummary %>%
  tidyr::separate(region_HIF, into = c("region", "HIF"),
                  sep = " ")
mySummary$HIF <- as.numeric(mySummary$HIF)
```

```{r, fig.cap="Indicator values (area weighted averages) for each region and homogeneous area class, as well as the number of data points per groups."}
ggarrange(
ggplot(mySummary, aes(x = HIF, y = area_weighted_mean_indicator_value,
                      fill = area_weighted_mean_indicator_value))+
  geom_bar(stat = "identity")+
  theme_bw(base_size = 12)+
  facet_wrap(.~region, ncol=1)+
  guides(fill = "none"),
ggplot(mySummary, aes(x = HIF, y = data_points,
                      fill = region))+
  geom_bar(stat = "identity")+
  theme_bw(base_size = 12)+
  facet_wrap(.~region, ncol=1)+
  guides(fill = "none")
)
```

The figure above shows that there are much more data on the two middle
classes (class 1 and 2). We do not have enough data to extrapolate to the 
highest impacted areas in any of the regions. These are not the areas prioritised for nature type mapping.

We also see that, even though there are small differences, high HIF is assiciated with lower indicator values.

```{r}
DT::datatable(mySummary)
```

```{r, eval=F}
wetlands_slitasje_extr <- ea_spread(indicator_data = wetlands,
                         indicator = i_2020,
                         regions = HIA_regions,
                         groups = region_HIF,
                         threshold = 100)
#saveRDS(wetlands_slitasje_extr, "P:/41201785_okologisk_tilstand_2022_2023/data/cache/wetlands_slitasje_extr.rds")
```
```{r, include=F}
wetlands_slitasje_extr <- readRDS("P:/41201785_okologisk_tilstand_2022_2023/data/cache/wetlands_slitasje_extr.rds")
```

@@

```{r}
wetlands_slitasje_extr_trd <- sf::st_crop(wetlands_slitasje_extr, myBB)
```

```{r, fig.cap="Sliatsje (wetland) indicator extrapolated over Trondheim"}
tm_shape(wetlands_slitasje_extr_trd)+
  tm_polygons(col = "area_weighted_mean_indicator_value",
    title="Slitasje indicator",
    palette = "viridis",
    style="cont",
    breaks = c(0,1))+
  tm_layout(legend.outside = T)+
  tm_shape(hw_utm)+
  tm_lines(col="blue")
```

Let's split the ID column to get the HIF and plot agains indicator values

```{r}
wetlands_slitasje_extr <- wetlands_slitasje_extr %>%
  tidyr::separate(ID, into = c("region", "HIF"),
                  sep = " ")
wetlands_slitasje_extr$HIF <- as.numeric(wetlands_slitasje_extr$HIF)
```

```{r}
summary(wetlands_slitasje_extr)
ggplot(wetlands_slitasje_extr, aes(x = factor(HIF), y = area_weighted_mean_indicator_value))+
         geom_boxplot(fill = "grey")+
  theme_bw(base_size = 12)
```

This relationship is quite good, and I think it justified using HIF as
homogeneous areas to extrapolate indicator values onto.

Split one more time, into regions:

```{r, eval=F}
wetlands_slitasje_extr2 <- ea_spread_dev(indicator_data = wetlands_slitasje_extr,
                         indicator = area_weighted_mean_indicator_value,
                         regions = regions,
                         groups = region,
                         summarise = F)
saveRDS(wetlands_slitasje_extr2, "P:/41201785_okologisk_tilstand_2022_2023/data/cache/wetlands_slitasje_extr2.rds")
```

```{r, include=F}
# silently read cached file
wetlands_slitasje_extr2 <- readRDS("P:/41201785_okologisk_tilstand_2022_2023/data/cache/wetlands_slitasje_extr2.rds")
```

```{r}
plot(factor(wetlands_slitasje_extr2$ID), wetlands_slitasje_extr2$area_weighted_mean_indicator_value,
     ylab="indicator value", xlab="", ylim=c(0,1))
```

>>>>

This rather dense figure needs to be looked at carefully. 
In the top row we see that the sample size (number of polygons) varies between ecosystems. For naturally open ecosystems many polygons are from high infrastructure areas, not surprisingly, and there are few points for the lowest impact class (zero such polygins in _Sørlandet_ for example). We there may not be able to extrapolate indicator values in areas in _Sørlandet_ with low levels of infrastructure. We can also from the middle row see the association between the indicator values (area weighted means) and the modified infrastructure index (HIF). We should then not put too much weight on data were there is little data (low n) and/or high standard error (sd, bottom row). For example, with Naturally open ecosystems, Midt-Norge has the overall lowest errors, and there we also see the kind of relationship between the indicator and the pressure that we want to see (middle row). The pattern is repeated for all regions except Østlandet, but there we also see that HIF class 3 has very little data, and a high sd, and this could explain the discrepancy (but it doesn't support the general relationship either).

@@

```{r, fig.cap="Summary statistics for the slitasje indicator."}
DT::datatable(all_stats)
```

@@

We can also calculate Kendall tau b coefficients, although I'm not complete sure this is valid when we have ordinal data with only four levels, since there will be a lot of ties.

```{r}
corrCheck_grouped <- as.data.frame(corrCheck) %>%
  group_by(hovedøkosystem, region) %>%
  summarise(cor = cor(i_2020, infrastructureIndex, method = "kendall"))
```

@@

```{r, fig.cap="A closer look at the HIA designation over Trondheim"}
(HIA_trd <- tm_shape(HIA_trd)+
  tm_polygons(col = "infrastructureIndex",
    title="Infrastructure index\n(modified 4-step scale)",
    palette = "-viridis",
    style="cat")+
  tm_layout(legend.outside = T)+
  tm_shape(hw_utm)+
  tm_lines(col="red")+
  tm_shape(outline)+
  tm_borders(col = "black", lwd=2))
```

@@

I want to add a threshold so that we don't end up over extrapolating
based on too few data points. From Fig. \@ref(fig:dense-stats) I conclude that we need at least 70 data points for the sd to stabilize. I will conservatively round this up to 100 data points.

```{r, eval=F}
wetlands_slitasje_extr <- ea_spread(indicator_data = wetlands,
                         indicator = i_2020,
                         regions = HIA_regions,
                         groups = region_HIF,
                         threshold = 100)

seminat_slitasje_extr <- ea_spread(indicator_data = seminat,
                         indicator = i_2020,
                         regions = HIA_regions,
                         groups = region_HIF,
                         threshold = 100)

natOpen_slitasje_extr <- ea_spread(indicator_data = natOpen,
                         indicator = i_2020,
                         regions = HIA_regions,
                         groups = region_HIF,
                         threshold = 100)
#saveRDS(wetlands_slitasje_extr, "P:/41201785_okologisk_tilstand_2022_2023/data/cache/wetlands_slitasje_extr.rds")
#saveRDS(seminat_slitasje_extr, "P:/41201785_okologisk_tilstand_2022_2023/data/cache/seminat_slitasje_extr.rds")
#saveRDS(natOpen_slitasje_extr, "P:/41201785_okologisk_tilstand_2022_2023/data/cache/natOpen_slitasje_extr.rds")
```
```{r, include=F}
wetlands_slitasje_extr <- readRDS("P:/41201785_okologisk_tilstand_2022_2023/data/cache/wetlands_slitasje_extr.rds")
seminat_slitasje_extr <- readRDS("P:/41201785_okologisk_tilstand_2022_2023/data/cache/seminat_slitasje_extr.rds")
natOpen_slitasje_extr <- readRDS("P:/41201785_okologisk_tilstand_2022_2023/data/cache/natOpen_slitasje_extr.rds")
```

@@

```{r}
wetlands_slitasje_extr_trd <- sf::st_crop(wetlands_slitasje_extr, myBB)
seminat_slitasje_extr_trd <- sf::st_crop(seminat_slitasje_extr, myBB)
natOpen_slitasje_extr_trd <- sf::st_crop(natOpen_slitasje_extr, myBB)
```

```{r, fig.cap="Slitasje indicator extrapolated over Trondheim"}
tmap_arrange(
tm_shape(wetlands_slitasje_extr_trd)+
  tm_polygons(col = "w_mean",
    title="Slitasje indicator",
    palette = "viridis",
    style="cont",
    breaks = c(0,1))+
  tm_layout(legend.outside = T)+
  tm_shape(hw_utm)+
  tm_lines(col="blue")+
  tm_layout(title = "Wetland"),

tm_shape(seminat_slitasje_extr_trd)+
  tm_polygons(col = "w_mean",
    title="Slitasje indicator",
    palette = "viridis",
    style="cont",
    breaks = c(0,1))+
  tm_layout(legend.outside = T)+
  tm_shape(hw_utm)+
  tm_lines(col="blue")+
  tm_layout(title = "Semi-natural areas"),

tm_shape(natOpen_slitasje_extr_trd)+
  tm_polygons(col = "w_mean",
    title="Slitasje indicator",
    palette = "viridis",
    style="cont",
    breaks = c(0,1))+
  tm_layout(legend.outside = T)+
  tm_shape(hw_utm)+
  tm_lines(col="blue")+
  tm_layout(title = "Naturally open areas"),
HIA_trd)
```

There is very little variation both between the ecosystems and the four classes in the modified infrastructure index. Note however, that for wetland and semi-natural areas we did not extrapolate values for the most urban areas, and that for naturally open areas we did not extrapolate to the least affected areas.

The next steps now are to

1. crop the extrapolated indicator maps to ecosystem delineation maps, and then;

2. aggregate these for each region or accounting area.
